---
title: "ex_11"
author: "Nikolai German"
date: "`r Sys.Date()`"
output: html_document
---

# Ex 1: Distributions

## a)

### i) Geometric Distribution

#### 2)

We take the Geometric Distribution quite literally, and just "flip" coins until we get a success, while counting, how many times we've been unsuccesfull.

```{r}
rgeom_custom <- function(n = 1, p = .5) {
  inner <- function() {
    x <- 0
    repeat {
      u <- runif(1)
      if (u < p) {
        break
      } else {
      x <- x + 1
      }
    }
  return(x)
  }
  replicate(n, inner())
}

samp <- rgeom_custom(100, .2)
hist(samp)
```


### ii) Cauchy Distribution

#### 2)
We use somewhat a reverse Box-Muller approach. We know that if we sample $r \in [0,1]$ and $\theta \in [0, 2\pi]$, we obtain points within the unit circle $(r \cos(\theta), r \sin(\theta))$. Thus $X/Y$ simplifies to $\frac{r \cos(\theta)}{r \sin(\theta)} = \frac{1}{\tan(\theta)}$

```{r}
set.seed(121)
samp <- 1/tan(2*pi*runif(100))
dens <- density(samp)

plot(dcauchy, xlim = c(-10,10), lty = "dashed")
lines(dens$x, dens$y, col = "blue")
```
### iii) Poisson Distribution

We simply sample independently from the exponential distribution until the cumulative sum is greater than 1. The number of samples, which did not cause to overflow is possion distributed.

```{r}

rpois_custom <- function(n, lambda = 1) {
  inner <- function() {
    t <- 0
    x <- 0
    while (t < 1) {
      t <- t - log(runif(1)) / lambda # Inversion for EXP
      if (t < 1) {
        x <- x + 1
      }
    }
    return(x)
  }
  replicate(n, inner())
}

samp <- rpois_custom(10000, 2)

hist(samp, freq = FALSE, breaks = 10)
plot(function(x) dpois(round(x), 2), add = TRUE, col = "blue", lwd = 3, xlim = c(0, 10))

```




## b)

The sum of squared i.i.d. standard normals is chi-squared distributed.

We use box-muller to generate i.i.d. standard normals:

```{r}
rchi2 <- function(n) {
  inner <- function() {
    r <- sqrt(-2*log(runif(1)))
    theta <- 2*pi*runif(1)
    X <- r * cos(theta)
    Y <- r * sin(theta)
    return(X^2 + Y^2)
  }
  replicate(n, inner())
}


samp <- rchi2(1000)
hist(samp, freq = FALSE, xlim = c(0,10))
plot(function(x) dchisq(x, df = 2), xlim = c(0,10), add = TRUE, lty = "dashed", lwd = 1.5)
```

```{r}
# other topic: worst case convergence of qmc

mc <- function(x) 1/sqrt(x)
qmc <- function(x, d = 1) ((log(x)^d)/x)

par(mfrow = c(2,3))
for (d in c(1, 3, 5, 10, 50, 100)) {
  plot(function(x) qmc(x, d=d), xlim = c(1,1e5), main = sprintf("d = %d", d), log = "xy")
  plot(mc, xlim = c(1,1e5), lty = "dashed", add = TRUE)
}


```

